---
title: "Metadata Label Transfer on Real Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Metadata Label Transfer on Real Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# Introduction

Here we take data from the [EiaD](eyeIntegration.nei.nih.gov) resource, which has been cut down 
to a few dozen cornea, retina, and RPE samples as well as the first few thousand
most variable genes (as calculated by variance). 

```{r, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE,  warning = FALSE,
  collapse = TRUE,
  fig.width = 5, fig.height = 3,
  comment = "#>"
)
```

# Load libraries
```{r}
library(projectR)
library(metamoRph)
library(dplyr)
library(ggplot2)
```

# Import EiaD data

An ocular subset of the full resource (full data can be found at eyeIntegration.nei.nih.gov)
```{r}

feature_by_sample <- data.table::fread(
  system.file('test_data/EiaD__eye_samples_counts.csv.gz', package = 'metamoRph')
)
genes <- feature_by_sample$Gene
feature_by_sample <- feature_by_sample[,-1] %>% as.matrix()
row.names(feature_by_sample) <- genes
meta_ref <- data.table::fread(
  system.file('test_data/EiaD__eye_samples_metaRef.csv.gz', package = 'metamoRph')
)
meta_project <- data.table::fread(
   system.file('test_data/EiaD__eye_samples_metaProject.csv.gz', package = 'metamoRph')
)
```

# projectR

As projectR does not apply the prcomp row (gene) scaling, the new data (red circle) ends up in the center of the PCA plot. Scaling on the new data alone also results in the variation between the samples being exaggerated.

```{r projectR_based_pca, fig.width=8, fig.height=5}
ref_matrix <- feature_by_sample[,meta_ref %>% pull(sample_accession)]
Pvars <- matrixStats::rowVars(log1p(ref_matrix))
select <- order(Pvars, decreasing = TRUE)[seq_len(min(1000,
                                                      length(Pvars)))]

pr_run <- prcomp(log1p(t(ref_matrix[select,])),
                 scale = TRUE,
                 center = TRUE)

new_mat_for_projection <- feature_by_sample[select,
                                            meta_project %>% pull(sample_accession)]

pr_proj <- projectR(data = log1p(as.matrix(new_mat_for_projection)),
                    loadings = pr_run,
                    dataNames = row.names(new_mat_for_projection))

data_plot <- bind_rows(cbind(t(pr_proj), meta_project %>%
                               mutate(Tissue = paste0(Tissue, " Projected"), Data = 'Projection')),
                       cbind(pr_run$x, meta_ref))
data_plot %>%
  mutate(Tissue = case_when(Cohort == 'Body' ~ 'Body', TRUE ~ Tissue)) %>%
  ggplot(aes(x=PC1,y=PC2, color = Tissue)) + geom_point() +
  ggforce::geom_mark_ellipse(data = data_plot %>% filter(Data == 'Projection'), color = 'red', size = 2) +
  scale_color_manual(values = pals::alphabet(n=12)%>% unname()) +
  ggforce::geom_mark_ellipse() +
  cowplot::theme_cowplot()

```

# metamoRph

metamoRph's two steps (`run_pca` and `metamoRph`) gets the new data into the proper scale for the projection
to work properly.
```{r metamoRph_eiad_pca, fig.width=8, fig.height=5}
pca_output <- run_pca(feature_by_sample = ref_matrix[select,],
                      meta = meta_ref,
                      hvg_selection = 'classic',
                      sample_cpm_scale = FALSE)

# project data from the pca
rownames(new_mat_for_projection) <- gsub(' \\(.*','',rownames(new_mat_for_projection))
projected_data <- metamoRph(new_mat_for_projection,
                            pca_output$PCA$rotation,
                            center_scale = pca_output$center_scale,
                            sample_cpm_scale =  FALSE)
data_plot <- bind_rows(cbind(projected_data, meta_project %>%
                               mutate(Tissue = paste0(Tissue, " Projected"),
                                      Data = 'Projection')),
                       cbind(pca_output$PCA$x, meta_ref) %>%
                         mutate(Data = 'Original'))
data_plot %>%
  mutate(Tissue = case_when(Cohort == 'Body' ~ 'Body', TRUE ~ Tissue)) %>%
  ggplot(aes(x=PC1,y=PC2,color = Tissue)) +
  geom_point() +
  scale_color_manual(values = pals::alphabet(n=12) %>% unname()) +
  ggforce::geom_mark_ellipse(data = data_plot %>% filter(Data == 'Projection'), color = 'red', size = 2) +
  ggforce::geom_mark_ellipse() +
  
  cowplot::theme_cowplot()
```

# metamoRph with cpm scaling and scran HVG selection

The CPM and scran steps modestly improves the PC1/PC2 distinction between the tissue types as 
well as pulling some outliers back towards their matching tissue type.
```{r metamoRph_eiad_pca_with_sample_scaling, fig.width=8, fig.height=5}
pca_output <- run_pca(ref_matrix[select,],
                      meta_ref,
                      hvg_selection = 'scran',
                      sample_cpm_scale = TRUE)

# project data from the pca
projected_data <- metamoRph(new_mat_for_projection,
                            pca_output$PCA$rotation,
                            center_scale = pca_output$center_scale,
                            sample_cpm_scale = TRUE)

data_plot <- bind_rows(cbind(projected_data, meta_project %>%
                               mutate(Tissue = paste0(Tissue, " Projected"),
                                      Data = 'Projection')),
                       cbind(pca_output$PCA$x, meta_ref) %>%
                         mutate(Data = 'Original'))
data_plot %>%
  mutate(Tissue = case_when(Cohort == 'Body' ~ 'Body', TRUE ~ Tissue)) %>%
  ggplot(aes(x=PC1,y=PC2,color = Tissue)) +
  geom_point() +
  scale_color_manual(values = pals::alphabet(n=12) %>% unname()) +
  ggforce::geom_mark_ellipse(data = data_plot %>% filter(Data == 'Projection'), color = 'red', size = 2) +
    ggforce::geom_mark_ellipse() +
  cowplot::theme_cowplot()
```

# Label Transfer
Finally we demonstrate the two step process for transferring over the labels from the input 
data onto the projected data.

First we run `model_build` on the original `pca_output` eigenvalue matrix (`$x`). You also have
to provide the label data of interest to transfer (`tissue`). Behind the scenes
a `lm` model is built for each unique label (in this case `tissue`) against all of the other tissues using the first 10 PCs. The output is a list of models (one for each tissue type).

We can then use this list of models in `model_apply` as well as the projected data from `metamoRph`. We provide (this is optional) the true labels of the projected data. A tibble is returned which gives the original label (if provided) in the `sample_label` field as well as the predicted label (the `predict`) field. The `max_score` is the confidence that the model has in the prediction. Closer to 0 is low confidence while closer to 1 is high confidence.
```{r}
trained_model <- model_build(pca_output$PCA$x,
                             pca_output$meta$Tissue,
                             model = 'lm', num_PCs = 10, verbose = FALSE,
                             BPPARAM = MulticoreParam(10))

label_guesses <- model_apply(trained_model,
                             projected_data,
                             meta_project$Tissue
)

label_guesses
```
