---
title: "Example of label projection"
output: html_notebook
---

# Acquire the eigenProjectR pre-made list objects
From [INSERT IN URL]
```{r}
#load('~/Downloads/scEiaD_seacell_pca_objs.Rdata')
# load('inst/data/scEiaD_seacell_pca_objs.Rdata')
```

# Acquire outside data
https://singlecell.broadinstitute.org/single_cell/study/SCP1755/joint-scrna-seq-and-scatac-seq-atlas-of-the-adult-human-retina#study-download

Download:
  - processed_barcodes.tsv.gz 
  - processed_features.tsv.gz 
  - processed_matrix.mtx.gz 
  
Make folder "SCP1755"

Put those three files into SCP1755 and hand delete the "processed_" part from each of the three names.

Also Download:
  - scPortal_metadata_file.txt 

# Example of how SCP755 should look (I did this all in my ~/Downloads folder)
```{r}
list.files('~/Downloads/test')
```


# Load in SCP755 
After loading the counts in as a `SingleCellExperiment` object, we then add in the cell type assignments to each cell
```{r}
library(tidyverse)
library(scran)
library(metamoRph)
load('../data/scEiaD_seacell_pca_objs_2023_07_10.Rdata')
# https://singlecell.broadinstitute.org/single_cell/study/SCP1755/joint-scrna-seq-and-scatac-seq-atlas-of-the-adult-human-retina
# download: processed_barcodes.tsv.gz; processed_features.tsv.gz; processed_matrix.tsv.gz; scPortal_metadata_file.txt
# hand delete the prefix "processed_" from each file
# place all four files in a folder of your choice
scp1755 <- DropletUtils::read10xCounts('~/Downloads/test/')
scp1755_meta <- read_tsv('~/Downloads/test/scPortal_metadata_file.txt')

scp1755_meta <- scp1755_meta[2:nrow(scp1755_meta),]

colData(scp1755)$NAME <- colnames(assay(scp1755))

colData(scp1755)$celltype <- colData(scp1755) %>% as_tibble() %>% mutate(NAME = Barcode) %>% left_join(scp1755_meta) %>% pull(cell_type__custom)

set.seed(123)
# splits each celltype randomly into 4 groups
scp1755_sum_counts <- aggregateAcrossCells(scp1755, paste(colData(scp1755)$celltype, round(runif(20,1,3), 0))) %>% assay()
#scp1755_sum_counts <- aggregateAcrossCells(scp1755, colData(scp1755)$celltype) %>% assay()
row.names(scp1755_sum_counts) <- rowData(scp1755)$Symbol

lr_model <- model_build(seacell_pca_human_adult$PCA$x,
                        seacell_pca_human_adult$meta$CellType,
                        verbose = FALSE, num_PCs = 50,
                        model = 'lm')

scp1755_projected <- metamoRph(scp1755_sum_counts,
                               rotation = seacell_pca_human_adult$PCA$rotation,
                               center_scale = seacell_pca_human_adult$center_scale)

model_apply(lr_model, scp1755_projected) %>% data.frame() %>% dplyr::select(sample_id, predict, max_score)

# scoring <- model_apply(lr_model, scp1755_projected, return_predictions = TRUE)
# bind_cols(model_apply(lr_model, scp1755_projected) %>% dplyr::select(sample_id) %>% as_tibble(),
#           scoring)

# ########################
custom.settings = umap::umap.defaults
custom.settings$n_neighbors = 30
custom.settings$min_dist <- 0.5
seacell_umap <- umap::umap(rbind(seacell_pca_human_adult$PCA$x,
                                 scp1755_projected)[,1:50],
                           config = custom.settings)
p <- seacell_umap$layout %>% as_tibble(rownames = 'info') %>%
  left_join(meta_mat %>% ungroup() %>% dplyr::select(seacell_id, study_accession, CellType, CTP), by = c('info' = 'seacell_id')) %>%
  mutate(label = paste0(info, '\n', study_accession, '\n',CTP)) %>%
  ggplot(aes(x=V1,y=V2,color=CellType, label = label)) +
  geom_point(size = 0.5) +
  cowplot::theme_cowplot() +
  scale_color_manual(values = c(pals::glasbey(), pals::alphabet2(), pals::alphabet2()) %>% unname())
plotly::ggplotly(p)

# ################
#harmony

# harmony_test <- harmony::HarmonyMatrix(rbind(scp1755_projected,
#                                              seacell_pca_human_adult$PCA$x)[,1:50],
#                                        do_pca = FALSE,
#                                        meta_data = c(rep('scp1755', nrow(scp1755_projected)), seacell_pca_human_adult[[1]]$x %>% row.names() %>% gsub('__\\d+','',.)),
#                                        max.iter.harmony = 1)

batchelor_test <- batchelor::fastMNN(rbind(scp1755_projected,
                                           seacell_pca_human_adult$PCA$x)[,1:50] %>% as.matrix() %>% t(),
                                     batch = c(rep('scp1755', nrow(scp1755_projected)), seacell_pca_human_adult[[1]]$x %>% row.names() %>% gsub('__\\d+','',.)),
                                     d= NA
) %>% reducedDim()

seacell_umap <- umap::umap(batchelor_test[,1:50], config = custom.settings)
p <- seacell_umap$layout %>% as_tibble(rownames = 'info') %>%
  left_join(seacell_summarise, by = c('info' = 'seacell_id')) %>%
  mutate(label = paste0(info, '\n', study_accession, '\n',CTP)) %>%
  ggplot(aes(x=V1,y=V2,color=CellType, label = label)) +
  geom_point(size = 0.5) +
  cowplot::theme_cowplot() +
  scale_color_manual(values = c(pals::glasbey(), pals::alphabet2(), pals::alphabet2()) %>% unname())
plotly::ggplotly(p)


lr_model <- model_build(batchelor_test[(nrow(scp1755_projected)+1):nrow(batchelor_test),],
                        seacell_pca_human_adult$meta$CellType,
                        verbose = FALSE, num_PCs = 50,
                        model = 'lm')

model_apply(lr_model, data.frame(batchelor_test[1:nrow(scp1755_projected),])) %>% data.frame()

```


